A weakly structured stem for human origins in Africa

### 2023-02-15 paper.tex
Fix quote format
shaved words from abstract

\begin{abstract}
While it is now broadly accepted that \emph{Homo sapiens}
originated within Africa, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in prior divergence time estimates.
Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models.
We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, contemporary
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135ka, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral \emph{Homo}
populations connected by gene flow over
hundreds of thousands of years (i.e., a weakly structured stem). We find that
weakly structured stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar.
Despite genetic similarity between these populations, an inferred 1--4\%
of genetic differentiation among contemporary human populations can be
attributed to genetic drift between stem populations.
We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
\end{abstract}

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) from southern Africa moved
northward thereby encountering Stem 2, or Stem 2 moved from central/western
Africa southward into Stem 1S, then we could observe disproportionate
ancestry contributions from different stems in contemporary groups. We observed
two merger events. The first, between Stem 1S and Stem 2, resulted in the
formation of an ancestral Khoe-San population $\sim120$ka. The second event
$\sim100$ka between Stem 1E and Stem 2, resulted in the formation of the
ancestors of eastern/western Africans including the ancestors of
``Out-of-Africans''.
% The following sentence may need clarification
Reticulated models do not have a unique and 
well-defined "basal" human population divergence.
We suggest conceptualizing the events $120$ka as the time of most recent
shared ancestry among sampled populations.
However, interpreting population divergence times in population genetics is 
always challenging due
to the co-estimation of divergence time and subsequent migration;
methods assuming clean vs reticulated splits can infer different split dates
(Figure~\ref{fig:supp-rccr-data} and Figure~\ref{fig:supp-IM-misspecification}).
Therefore, wide variation exists in estimates of divergence time in the
literature\citep{Bergstrom2021-iw,Henn2018-rf}.

>>>>

\begin{abstract}
Despite broad agreement that \emph{Homo sapiens}
originated within Africa, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by limited fossil and genomic data, as
well as variability in prior divergence time estimates.
We use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models.
We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as new whole-genome sequence data for 44 Nama (Khoe-San) individuals. 
We infer a reticulated African population history in which contemporary
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135ka, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer structure between two or more weakly
differentiated ancestral \emph{Homo}
populations connected by gene flow over
hundreds of thousands of years (i.e., a weakly structured stem). Weakly structured stem 
models provide more likely explanations of patterns of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar.
Despite genetic similarity between these populations, an inferred 1--4\%
of genetic differentiation among contemporary human populations can be
attributed to genetic drift between stem populations.
We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
\end{abstract}

@@

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) from southern Africa moved
northward thereby encountering Stem 2, or Stem 2 moved from central/western
Africa southward into Stem 1S, then we could observe disproportionate
ancestry contributions from different stems in contemporary groups. We observed
two merger events. The first, between Stem 1S and Stem 2, resulted in the
formation of an ancestral Khoe-San population $\sim120$ka. The second event
$\sim100$ka between Stem 1E and Stem 2, resulted in the formation of the
ancestors of eastern/western Africans including the ancestors of
``Out-of-Africans''.
% The following sentence may need clarification
Reticulated models do not have a unique and 
well-defined ``basal'' human population divergence.
We suggest conceptualizing the events $120$ka as the time of most recent
shared ancestry among sampled populations.
However, interpreting population divergence times in population genetics is 
always challenging due
to the co-estimation of divergence time and subsequent migration;
methods assuming clean vs reticulated splits can infer different split dates
(Figure~\ref{fig:supp-rccr-data} and Figure~\ref{fig:supp-IM-misspecification}).
Therefore, wide variation exists in estimates of divergence time in the
literature\citep{Bergstrom2021-iw,Henn2018-rf}.
